---
title: "Midterm"
subtitle: "Biomass Visualization"
author: "Alexis Means"
date: 2025-03-26
description: I will be using quality and biomass data to help visualize how forage quality changes throughout my study area over time. 
code-fold: TRUE
---

# Project Description:

## **Understanding Bighorn Sheep Nutrition and Movement in the John Day Canyon**

Located within the Columbia Plateau, the John Day Canyon is a steep, rugged landscape where sagebrush and grasses dominate the hillsides. This dynamic environment supports a unique population of bighorn sheep — one that, remarkably, has avoided the disease outbreaks that often devastate other herds.

My research focuses on understanding the nutritional requirements of these bighorn sheep and how those needs shift throughout their reproductive cycle due to changing metabolic demands. The ultimate goal is to develop a predictive model that links their nutritional requirements to movement patterns. By doing so, we can identify potential threats, such as areas where bighorn sheep may come into contact with domestic livestock, and pinpoint critical habitats essential for their survival.

To achieve this, I am collecting vegetation data to calculate a measure of “**suitable biomass**”—essentially, the portion of available forage that meaningfully contributes to the sheep’s energy needs at different reproductive stages. This metric is created using measures of digestible energy and digestible protein. I have currently not finalized my values of "suitable biomass" but I could look at the preliminary values of digestible energy and digestible protein. Eventually using this metric, I will develop a Generalized Additive Model (GAM) to map areas of high and low nutritional quality across the landscape.

These techniques will can help determine which resources most influence bighorn sheep movement and allow us to predict future movement patterns—a critical step in ensuring their continued success in this landscape. By bridging nutritional ecology with movement ecology, this research will contribute valuable insights into how bighorn sheep navigate their environment and help inform conservation strategies for this unique, disease-free population.

### Data Wrangling

```{r}
#| warning: FALSE

library(lpSolve)
library(openxlsx)
library(tools)
library(dplyr)
library(sf)
library(tidyverse)

# Define the path to the directory containing the file
excelpath <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/FRESH/processed.data/"

# Define the specific file name
filename <- "FRESH.Subset.xlsx"

# Construct the full file path
excel.file <- file.path(excelpath, filename)

# Load the workbook and read the data
library(openxlsx)
habitat <- loadWorkbook(excel.file)
data <- data.frame(readWorkbook(excel.file, sheet = "FRESH Data", startRow = 1, colNames = TRUE))


plant <-read.csv("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/Cleaning/processed.data/plant.csv")

data <- data %>% 
  rename(Code = Plant.Code) %>% 
  left_join(y = plant, 
            by = "Code",
            relationship = "many-to-one")


```

## Exploring Quality Trends

### Comparing DE and DP values for phenological stage

With some of my quality data, I wanted to start by comparing the measurements of digestible energy (DE) and digestible protien (DP) between plant species and phenological stages that I observed while I was sampling. Mainly I want to observe if there is any correlation between high DE and high DP. After looking at the results of my scatter plots, the trend follows the patterns that I would expect. There is an increase in DE and DP through the newly emergent (N) and budding stages (B), followed by an immedaite decrease in the flowering (FL) stage.

```{r}
#| warning: FALSE
library(tidyverse)

data$Phenology <- factor(data$Phenology, levels = c("N", "B", "FL", "FR", "M", "C"))

ggplot(data, aes(x = DP, y = DE)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +  # Linear trend line
  labs(title = "Relationship Between DE and DP by Phenology",
       x = "Digestible Protein (DP)",
       y = "Digestible Energy (DE)") +
  theme_minimal() +
  
  # Set phenology as a factor with a specific order
  facet_wrap(~ Phenology, scales = "free") +
  scale_x_continuous(expand = c(0.05, 0)) +  # Adjust spacing
  scale_y_continuous(expand = c(0.05, 0)) +
  theme(strip.text = element_text(face = "bold"))  # Bold facet labels


```

To better visualize the trends in DP and DE between different phenological stages. I isolated the DE and DP and values to compare each state. Both the DE and DP values follow the trend that I expect them to, however, very few of my species meet the necessary DP value of 7.5.

```{r}
#| warning: FALSE
# Ensure Phenology is a factor with the correct order
data$Phenology <- factor(data$Phenology, levels = c("N", "B", "FL", "FR", "M", "C"))

ggplot(data, aes(x = Phenology, y = DE, fill = Phenology)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) by Phenology",
       x = "Phenology",
       y = "DE(kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none")+
   geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1)+
  annotate("text", x = 6.4, y = 11.8, label = "11.5", color = "black", size = 4, fontface = "italic")
```

```{r}
#| warning: FALSE
ggplot(data, aes(x = Phenology, y = DP, fill = Phenology)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  # Boxplot without outlier points
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +  # Add jitter for visibility
  scale_fill_viridis_d(option = "plasma") +  # Use a color palette
  labs(title = "Distribution of Digestible Protien (DP) by Phenology",
       x = "Phenology",
       y = "DP (g protien/100g forage)") +
  theme_minimal() +
  theme(legend.position = "none")+
   geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1)+
  annotate("text", x = 6.5, y = 7.7, label = "7.5", color = "black", size = 4, fontface = "italic")
```

### DE and DP values compared by PVT

```{r}
#| warning: FALSE
data <- data %>%
  mutate(PVT = str_sub(TransectID, start = 4, end = 6))

ggplot(data, aes(x = PVT, y = DE, fill = Phenology)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) by Phenology and PVT",
       x = "Phenology",
       y = "DE(kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5) +
  facet_wrap(~ Phenology, scales = "free_x", ncol = 3)  # Adjust ncol as needed

```

```{r}
#| warning: FALSE
# Filter the data for Phenology "N"
data_N <- data %>% filter(Phenology == "N")

ggplot(data_N, aes(x = PVT, y = DE, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) for Phenology 'N' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DE (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "B"
data_B <- data %>% filter(Phenology == "B")

ggplot(data_B, aes(x = PVT, y = DE, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) for Phenology 'B' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DE (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "B"
data_FL <- data %>% filter(Phenology == "FL")

ggplot(data_FL, aes(x = PVT, y = DE, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) for Phenology 'FL' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DE (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "FR"
data_FR <- data %>% filter(Phenology == "FR")

ggplot(data_FR, aes(x = PVT, y = DE, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) for Phenology 'FR' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DE (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "M"
data_M <- data %>% filter(Phenology == "M")

ggplot(data_M, aes(x = PVT, y = DE, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) for Phenology 'M' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DE (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "C"
data_C <- data %>% filter(Phenology == "C")

ggplot(data_C, aes(x = PVT, y = DE, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Energy (DE) for Phenology 'C' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DE (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 11.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 11.8, label = "11.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

```

```{r}
#| warning: FALSE

# Filter the data for Phenology "N"
data_N <- data %>% filter(Phenology == "N")

ggplot(data_N, aes(x = PVT, y = DP, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Protein (DP) for Phenology 'N' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DP (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 7.8, label = "7.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "B"
data_B <- data %>% filter(Phenology == "B")

ggplot(data_B, aes(x = PVT, y = DP, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Protein (DP) for Phenology 'B' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DP (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 7.8, label = "7.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "B"
data_FL <- data %>% filter(Phenology == "FL")

ggplot(data_FL, aes(x = PVT, y = DP, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Protein (DP) for Phenology 'FL' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DP (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 7.8, label = "7.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "FR"
data_FR <- data %>% filter(Phenology == "FR")

ggplot(data_FR, aes(x = PVT, y = DP, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Protein (DP) for Phenology 'FR' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DP (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 7.8, label = "7.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "M"
data_M <- data %>% filter(Phenology == "M")

ggplot(data_M, aes(x = PVT, y = DP, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Protein (DP)) for Phenology 'M' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DP (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 7.8, label = "7.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

# Filter the data for Phenology "C"
data_C <- data %>% filter(Phenology == "C")

ggplot(data_C, aes(x = PVT, y = DP, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +   
  scale_fill_viridis_d(option = "plasma") +  
  labs(title = "Distribution of Digestible Protien (DP) for Phenology 'C' by PVT",
       x = "Plant Vegetation Type (PVT)",
       y = "DP (kJ g^-1)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 7.5, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = Inf, y = 7.8, label = "7.5", color = "black", size = 4,
           hjust = 1.1, vjust = -0.5)

```

```{r}
plot <- read.csv("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/Cleaning/processed.data/transect.csv")

# Define the path to the directory containing the file
excelpath <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/FRESH/processed.data/"

# Define the specific file name
filename <- "2024.subset.results.xlsx"

# Construct the full file path
excel.file <- file.path(excelpath, filename)

# Load the workbook and read the data

habitat <- loadWorkbook(excel.file)
suitable <- data.frame(readWorkbook(excel.file, sheet = "Plot-Level-Summary", startRow = 1, colNames = TRUE))


suitable <- suitable %>% 
  rename(PlotID = TransectID) %>% 
  left_join(y = plot,
            by = "PlotID")

suitable <- suitable %>% 
  rename(Lat = BeginLat,
         Long = BeginLong,
         Easting = BeginUTM_Easting,
         Northing = BeginUTM_Northing) %>% 
  select(PlotID, TotalDE, TotalDP, SuitableBiomass, AveDE, AveDP, Julian_Day, Lat, Long, Easting, Northing)


# Create an sf object with the Lat and Long columns
suitable_UTM <- st_as_sf(suitable, coords = c("Long", "Lat"), crs = 4326)  # EPSG:4326 is WGS84

# Convert to UTM Zone 11N (EPSG:26911)
suitable_UTM <- st_transform(suitable_UTM, crs = 32611)

# Extract UTM coordinates (Easting and Northing)
suitable$Easting <- st_coordinates(suitable_UTM)[, 1]
suitable$Northing <- st_coordinates(suitable_UTM)[, 2]

```

```{r}
suitable <- suitable %>%
  mutate(PVT = str_sub(PlotID, start = 4, end = 6))

# Step 1: Standardize the "Suitable Biomass" values
suitable_standardized <- suitable %>%
  group_by(PVT) %>%
  mutate(Standardized_Biomass = scale(SuitableBiomass)) %>%
  ungroup()

# Step 2: Create the plot
ggplot(suitable_standardized, aes(x = PVT, y = Standardized_Biomass, fill = PVT)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Comparison of Standardized Suitable Biomass Across PVTs",
       x = "Plant Vegetation Type (PVT)",
       y = "Standardized Suitable Biomass") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) 
```

### Quality Trends overtime 

```{r}
#| warning: FALSE
ggplot(suitable, aes(x = Julian_Day, y = SuitableBiomass)) +
  geom_point(aes(color = PVT), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  scale_color_viridis_d() +
  labs(title = "Suitable Biomass Over Time",
       x = "Julian Day",
       y = "Suitable Biomass",
       color = "PVT") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  scale_x_continuous(breaks = seq(min(suitable$Julian_Day, na.rm = TRUE), 
                                  max(suitable$Julian_Day, na.rm = TRUE), 
                                  by = 30))

```

```{r}
#| warning: FALSE
ggplot(suitable, aes(x = Julian_Day, y = AveDE)) +
  geom_point(aes(color = PVT), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  scale_color_viridis_d() +
  labs(title = "Average DE Over Time",
       x = "Julian Day",
       y = "DE",
       color = "PVT") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  scale_x_continuous(breaks = seq(min(suitable$Julian_Day, na.rm = TRUE), 
                                  max(suitable$Julian_Day, na.rm = TRUE), 
                                  by = 30))


```

```{r}
#| warning: FALSE
ggplot(suitable, aes(x = Julian_Day, y = AveDP)) +
  geom_point(aes(color = PVT), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  scale_color_viridis_d() +
  labs(title = "Average DP Over Time",
       x = "Julian Day",
       y = "DP",
       color = "PVT") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  scale_x_continuous(breaks = seq(min(suitable$Julian_Day, na.rm = TRUE), 
                                  max(suitable$Julian_Day, na.rm = TRUE), 
                                  by = 30))


```

## Biomass Trends

One of the key parts of my project is understanding the spread of invasive species throughout the canyon, specifically cheat grass. When observing the canyon it seems like cheat grass dominates the canyon which could effect the amount of nutrients available later in the summer. To start I wanted to looks at the density of biomass for all native vs invasive species within the transects that I sampled in our study area. From this graph we can see that overall biomass of native and invasive species is relatively similar throughout the canyon. However, there is an overall greater density of native species.

```{r}
spp.map <- data %>% 
  select(TransectID, Code, Phenology, Part, Biomass, Duration, Status)

plot <- read.csv("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/Cleaning/processed.data/transect.csv")

spp.map <- spp.map %>% 
  rename(PlotID = TransectID)

spp.map <- spp.map %>% 
  left_join(y = plot,
            by = "PlotID",
            relationship = "many-to-one")

spp.map <- spp.map %>% 
  rename(Lat = BeginLat,
         Long = BeginLong,
         Easting = BeginUTM_Easting,
         Northing = BeginUTM_Northing) %>% 
  select(PlotID, Code, Phenology, Part, Biomass, Duration, Status, Julian_Day, Lat, Long, Easting, Northing)


# Create an sf object with the Lat and Long columns
spp.map_UTM <- st_as_sf(spp.map, coords = c("Long", "Lat"), crs = 4326)  # EPSG:4326 is WGS84

# Convert to UTM Zone 11N (EPSG:26911)
spp.map_UTM <- st_transform(spp.map_UTM, crs = 32611)

# Extract UTM coordinates (Easting and Northing)
spp.map$Easting <- st_coordinates(spp.map_UTM)[, 1]
spp.map$Northing <- st_coordinates(spp.map_UTM)[, 2]
```

```{r}
#| warning: FALSE
# Load necessary libraries
library(tidyverse)

# Summarize biomass weight by spatial location and species status
spp.summary <- spp.map %>%
   filter(!is.na(Biomass) & Biomass > 0)  # Remove NAs
  
ggplot(spp.summary, aes(x = Easting, y = Northing, weight = Biomass)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  facet_wrap(~ Status) +  # Separate maps for Native/Invasive
  scale_fill_viridis_c(
    name = "Biomass Density", 
    option = "turbo",
    breaks = c(1e-09, 7e-09),  # Define specific density values
    labels = c("Low", "High")  # Replace numerical values with "Low" and "High"
  ) +  
  labs(title = "Density of Biomass (Native vs Invasive Species)",
       x = "Easting", 
       y = "Northing") +
  theme_minimal() +
  theme(axis.text = element_blank(),  # Remove tick labels
    axis.ticks = element_blank())  # Rotate x-axis labels

```

I still do not have complete quality data, so there is only so much I can do with this information, but knowing the amount of biomass for each species in the area will allow us to apply a nutrient value of "suitable biomass". I would also like to see the differences in overall biomass for each month. I had a much larger sample size in April and May than I did in June and July, so I standardized my biomass values before plotting them.

```{r}
#| warning: FALSE
month.sum <- spp.map %>%
  filter(!is.na(Biomass) & Biomass > 0) %>%  # Remove NAs
  mutate(Month = case_when(
    Julian_Day >= 93 & Julian_Day < 124 ~ "April",
    Julian_Day >= 124 & Julian_Day < 155 ~ "May",
    Julian_Day >= 155 & Julian_Day < 185 ~ "June",
    Julian_Day >= 185 & Julian_Day <= 212 ~ "July",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Month)) %>%
  mutate(Month = factor(Month, levels = c("April", "May", "June", "July")))

# Normalize biomass within each month (to make sample sizes comparable)
month.sum <- month.sum %>%
  group_by(Month) %>%
  mutate(Total_Biomass = sum(Biomass),  # Calculate total biomass for each month
         Normalized_Biomass = Biomass / Total_Biomass) %>%  # Normalize biomass
  ungroup()  # Remove grouping


# Create density plot faceted by Month and Status
ggplot(month.sum, aes(x = Easting, y = Northing, weight = Normalized_Biomass)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  facet_grid(Month ~ Status) +  # Grid facet by Month (rows) and Status (columns)
  scale_fill_viridis_c(
    name = "Biomass Density", 
    option = "turbo",
    breaks = c(2e-09, 18e-09),  # Define specific density values
    labels = c("Low", "High")  # Replace numerical values with "Low" and "High"
  ) +  
  labs(title = "Normalized Density of Biomass (Native vs Invasive Species by Month)",
       x = "Easting", 
       y = "Northing") +
  theme_minimal() +
  theme(axis.text = element_blank(),  # Remove tick labels
        axis.ticks = element_blank())
```

Since cheat grass is the main issue that we have been observing as far as the most invasive species in the canyon I would also like to see how this biomass progresses over each month. Apparently in this subset of my data I didn't have any observations for cheatgrass in the month of June. This makes me think that I should take the rest of my results with a grain of salt until I can complete my biomass predictions. That being said the density of biomass for cheat grass did peak in the southern end of the canyon in June and then started to decrease after that point.

```{r}
#| warning: FALSE
# Filter for observations with Species Code "BRTE"
cheat.sum <- spp.map %>%
  filter(!is.na(Biomass) & Biomass > 0, Code == "BRTE") %>%  # Remove NAs and filter for "BRTE"
  mutate(Month = case_when(
    Julian_Day >= 93 & Julian_Day < 124 ~ "April",
    Julian_Day >= 124 & Julian_Day < 155 ~ "May",
    Julian_Day >= 155 & Julian_Day < 185 ~ "June",
    Julian_Day >= 185 & Julian_Day <= 212 ~ "July",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Month)) %>%
  mutate(Month = factor(Month, levels = c("April", "May", "June", "July")))  # Set correct order

# Normalize biomass within each month (to make sample sizes comparable)
cheat.sum <- cheat.sum %>%
  group_by(Month) %>%
  mutate(Total_Biomass = sum(Biomass),  # Calculate total biomass for each month
         Normalized_Biomass = Biomass / Total_Biomass) %>%  # Normalize biomass
  ungroup()  # Remove grouping

# Create density plot faceted by Month and Status for species "BRTE"
ggplot(cheat.sum, aes(x = Easting, y = Northing, weight = Normalized_Biomass)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  facet_grid(Month ~ Code) +  # Grid facet by Month (rows) and Status (columns)
  scale_fill_viridis_c(
    name = "Biomass Density", 
    option = "turbo",
    breaks = c(2e-09, 14e-09),  # Define specific density values
    labels = c("Low ", "High ")  # Replace numerical values with "Low" and "High"
  ) +  
  labs(title = "Normalized Density of Biomass for BRTE by month",
       x = "Easting", 
       y = "Northing") +
  theme_minimal() +
  theme(axis.text = element_blank(),  # Remove tick labels
        axis.ticks = element_blank())
```

## GPS Data

Lastly, I want to see how the sheep movements change each month with the overall biomass of both native and invasive species. I started out by creating a heat map of our overall sheep location use throughout the canyon.

```{r}
#| warning: FALSE
gps <- read.csv("C:/Users/Alexis Means/Documents/Project/Demographics/Code/cleaned.data/sheep_clean_24.csv")

gps <- gps %>% 
  rename(Easting = x_,
         Northing = y_)

# Step 1: Kernel Density Estimation (KDE) for GPS data
gps_kde <- ggplot(gps, aes(x = Easting, y = Northing)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "inferno",
          name = "GPS Use Density",
          breaks = c(2e-09, 10e-09),  # Define specific density values
          labels = c("Low", "High")  # Replace  values with "Low" and "High"
  ) +
  labs(title = "Sheep Area Use")+
  theme_minimal()+
  theme(axis.text = element_blank(),  # Remove tick labels
        axis.ticks = element_blank()) 
  plot(gps_kde)
  
  
```

If the areas of high biomass density align with values of high nutrient quality in theory this means that the density of our sheep points should also follow the trends of high biomass density in the months of April and May. I wanted to see if I could overlay our sheep locations each month with the biomass density data each month to observe my hunch that sheep would follow this trend in these months. They do not appear to follow this trend, most likely due to other outside factors such as selection for escape terrain that outweigh the need to follow vegetation trends.

```{r}
#| warning: FALSE
month.sum <- spp.map %>%
  filter(!is.na(Biomass) & Biomass > 0) %>%  # Remove NAs
  mutate(Month = case_when(
    Julian_Day >= 93 & Julian_Day < 124 ~ "April",
    Julian_Day >= 124 & Julian_Day < 155 ~ "May",
    Julian_Day >= 155 & Julian_Day < 185 ~ "June",
    Julian_Day >= 185 & Julian_Day <= 212 ~ "July",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Month)) %>%
  mutate(Month = factor(Month, levels = c("April", "May", "June", "July")))


# Normalize biomass within each month (to make sample sizes comparable)
month.sum <- month.sum %>%
  group_by(Month) %>%
  mutate(Total_Biomass = sum(Biomass),  # Calculate total biomass for each month
         Normalized_Biomass = Biomass / Total_Biomass) %>%  # Normalize biomass
  ungroup()  # Remove grouping

# Extract month from the GPS timestamp
gps <- gps %>%
  mutate(Month = case_when(
    month(t_) == 4 ~ "April",
    month(t_) == 5 ~ "May",
    month(t_) == 6 ~ "June",
    month(t_) == 7 ~ "July",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Month)) %>% 
  mutate(Month = factor(Month, levels = c("April", "May", "June", "July")))

ggplot() + # Biomass density heatmap using kernel density estimation
  stat_density_2d(
    data = month.sum,
    aes(x = Easting, y = Northing, weight = Normalized_Biomass, fill = after_stat(density)),
    geom = "raster",
    contour = FALSE
  ) +
  facet_grid(Month ~ Status) +  # Grid facet by Month (rows) and Status (columns)
  scale_fill_viridis_c(
    name = "Biomass Density", 
    option = "turbo",
    breaks = c(2e-09, 18e-09),  # Define specific density values
    labels = c("Low", "High")  # Replace numerical values with "Low" and "High"
  ) +  
  labs(title = "Normalized Density of Biomass with GPS Locations (Native vs Invasive Species by Month)",
       x = "Easting", 
       y = "Northing") +
  theme_minimal() +
  theme(axis.text = element_blank(),  # Remove tick labels
        axis.ticks = element_blank()) +
  # Overlay GPS points, filtering by Month to match the facet
  geom_point(
    data = gps, 
    aes(x = Easting, y = Northing), 
    color = "red", alpha = 0.1, size = 0.4
  )
```

```{r}
# Define the path to the directory containing the file
excelpath <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/FRESH/processed.data/"

# Define the specific file name
filename <- "2024.subset.results.xlsx"

# Construct the full file path
excel.file <- file.path(excelpath, filename)

# Load the workbook and read the data

habitat <- loadWorkbook(excel.file)
suitable <- data.frame(readWorkbook(excel.file, sheet = "Plot-Level-Summary", startRow = 1, colNames = TRUE))


suitable.map <- suitable %>% 
  rename(PlotID = TransectID) %>% 
  left_join(y = plot,
            by = "PlotID")

suitable.map <- suitable.map %>% 
  rename(Lat = BeginLat,
         Long = BeginLong,
         Easting = BeginUTM_Easting,
         Northing = BeginUTM_Northing) %>% 
  select(PlotID, TotalDE, TotalDP, SuitableBiomass, AveDE, AveDP, Julian_Day, Lat, Long, Easting, Northing)
View(suitable.map)

# Create an sf object with the Lat and Long columns
suitable.map_UTM <- st_as_sf(suitable.map, coords = c("Long", "Lat"), crs = 4326)  # EPSG:4326 is WGS84

# Convert to UTM Zone 11N (EPSG:26911)
suitable.map_UTM <- st_transform(suitable.map_UTM, crs = 32611)

# Extract UTM coordinates (Easting and Northing)
suitable.map$Easting <- st_coordinates(suitable.map_UTM)[, 1]
suitable.map$Northing <- st_coordinates(suitable.map_UTM)[, 2]
```

## Suitable Biomass

After completing the FRESH model with synthetic values for crude protein using a subset of my data, I was able to get estimates of usable suitable biomass available at each transect that I sampled. I replicated similar graphs to my biomass density for native vs invasive species and the results are fairly similar.Like the biomass plots, there is a higher density in the south end of our study area when looking at the suitable biomass overall for the study area.

```{r}
#| warning: FALSE
suitable.summary <- suitable.map %>%
   filter(!is.na(SuitableBiomass) & SuitableBiomass > 0)  # Remove NAs
  
ggplot(suitable.summary, aes(x = Easting, y = Northing, weight = SuitableBiomass)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(
    name = "Suitable Biomass Density", 
    option = "turbo",
    breaks = c(0.5e-09, 3.5e-09),  # Define specific density values
    labels = c("Low", "High")  # Replace numerical values with "Low" and "High"
  ) +  
  labs(title = "Density of Suitable Biomass",
       x = "Easting", 
       y = "Northing") +
  theme_minimal() +
  theme(axis.text = element_blank(),  # Remove tick labels
    axis.ticks = element_blank())  # Rotate x-axis labels
```

```{r}
#|warning: FALSE
suitablemonth.sum <- suitable.map %>%
  filter(!is.na(SuitableBiomass) & SuitableBiomass > 0) %>%  # Remove NAs
  mutate(Month = case_when(
    Julian_Day >= 93 & Julian_Day < 124 ~ "April",
    Julian_Day >= 124 & Julian_Day < 155 ~ "May",
    Julian_Day >= 155 & Julian_Day < 185 ~ "June",
    Julian_Day >= 185 & Julian_Day <= 212 ~ "July",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Month)) %>%
  mutate(Month = factor(Month, levels = c("April", "May", "June", "July")))

# Normalize biomass within each month (to make sample sizes comparable)
suitablemonth.sum <- suitablemonth.sum %>%
  group_by(Month) %>%
  mutate(Total_SuitableBiomass = sum(SuitableBiomass),  # Calculate total biomass for each month
         Normalized_SuitableBiomass = SuitableBiomass / Total_SuitableBiomass) %>%  # Normalize biomass
  ungroup()  # Remove grouping


# Create density plot faceted by Month and Status
ggplot(suitablemonth.sum, aes(x = Easting, y = Northing, weight = Normalized_SuitableBiomass)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  facet_grid("Month") +  # Grid facet by Month (rows) and Status (columns)
  scale_fill_viridis_c(
    name = "Suitable Biomass Density", 
    option = "turbo",
    breaks = c(1e-09, 9e-09),  # Define specific density values
    labels = c("Low ", "High ")  # Replace numerical values with "Low" and "High"
  ) +  
  labs(title = "Normalized Density of Suitable Biomass",
       x = "Easting", 
       y = "Northing") +
  theme_minimal() +
  theme(axis.text = element_blank(),  # Remove tick labels
        axis.ticks = element_blank())
```

When looking at the density of suitable biomass month by month, May still has the highest density of suitable biomass. While this is most likely the month with the largest amount of species in their nutrient dense phase, it is also the month that we sampled the most. So I feel like the sampling efforts may be slightly skewing my data.

In order to truly see these values accurately represented across my study area, I will need to start creating my predictive GAM. This will allow me to project these values of suitable biomass to other 30x30m pixels with similar covariates and predict how vegetative quality changes throughout the canyon in different times of the year.
