---
title: "Suitable Biomass Subset"
author: "Alexis Means"
date: 2025-07-19
code-fold: TRUE
---

# **Suitable Biomass**

Using the subset of my data that I have created, I want to set up some visualizations that I know will be useful once I have the completed version of my data set. All of the values provided in my data set are actual measurements of samples. However I have not completed my linear regression equation that allows me to predict biomass for my unclipped plots. So for now this data represents just my clipped quadrats

## Data wrangling

```{r}

library(tidyverse) 
library(lubridate) 
library(sf)

setwd("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/FRESH/processed.data/") 
totals <- read.csv("test.totals.csv")
fdata <- read.csv("test.fdata.csv") 
transect <- read.csv("../raw.data/transect.csv") #formatting plot database 

plot <- transect %>% rename(TransectID = PlotID, Lat = BeginLat, Long = BeginLong) %>% select(Dates, TransectID, PVT, Lat, Long)

plot <- plot %>% 
  mutate(Dates = mdy(Dates), JulianDay = yday(Dates)) %>%
  st_as_sf(coords = c("Long", "Lat"), crs = 4326) %>% 
  st_transform(crs = 32611) %>% 
  mutate(Easting = st_coordinates(.)[,1], Northing = st_coordinates(.)[,2]) %>% select(JulianDay, TransectID, PVT, Easting, Northing) %>% st_drop_geometry()

#formatting fdata 
fdata <- fdata %>% select(-Max, -Pct_Used, -SuitableBiomass)

#combine all the databases 
data <- plot %>% 
  left_join(y = totals, by = "TransectID") %>% 
  left_join(y = fdata, by = "TransectID", relationship = "one-to-many") 

#I hate the way it is arranged, so I am going to rearrange it 
data <- data %>% 
  select(JulianDay, PVT, TransectID, Plant.Code, Phenology, Part, Biomass, Biomass_Used, Pct_Suitable_Biomass, SuitableBiomass, DE, TotalDE, AveDE, DP, TotalDP, AveDP, Easting, Northing) 

#recategorize PVT
data <- data %>%
  mutate(PVT = recode(PVT, `672` = "Grassland", `668` = "Scabland", `669` = "Sagebrush Shrubland", `674` = "Sagebrush Steppe", `682` = "Riparian"))

rm(fdata, plot, totals, transect)

```

## Suitable Biomass Visualizations

```{r}
# Calculate the percent of biomass relative to each PVT's total biomass percent
percentdata <- data %>% group_by(PVT) %>%
  mutate( total_biomass = sum(SuitableBiomass, na.rm = TRUE),  # Total biomass in each PVT 
          percent_biomass = SuitableBiomass / total_biomass * 100 # Percent biomass relative to total biomass in PVT
          ) %>% ungroup() 

# Boxplot for percent biomass 
ggplot(percentdata, aes(x = PVT, y = percent_biomass, fill = PVT)) +
  geom_boxplot() + 
  labs(title = "% Suitable Biomass Relative to Total Suitable Biomass by PVT", x = "PVT", y = "Percent Suitable Biomass") + 
  theme_minimal() + theme(legend.position = "none")
```

**DE Visualizations**

**DP Visualizations**

```{r}
# Create the boxplot
ggplot(data, aes(x = PVT, y = SuitableBiomass, fill = PVT)) + 
  geom_boxplot() + 
  labs(title = "% Suitable Biomass Relative to Total Suitable Biomass by PVT", x = "PVT", y = "Suitable Biomass per PVT") +
  theme_minimal() + 
  theme(legend.position = "none") 
```
