---
title: "Checking Biomass Predictions"
author: "Alexis Means"
date: "2025-10-2"
description: This document contains assumption tests and plots to analyze my linear regression equations for biomass predictions. 
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
editor: visual
code-fold: true
execute: 
  warning: false
  message: false
---

```{r}
spp_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Species-Top-Model-List.rds")

fam_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Family-Top-Model-List.rds")

genus_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Genus-Top-Model-List.rds")

fg_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Functional-Group-Top-Model-List.rds")

all_models <- list(
  spp   = spp_model,
  fam   = fam_model,
  genus = genus_model,
  fg    = fg_model
)

```

## Predicted Biomass 

This data frame includes all the predicted biomass measurements in a table if you are curious to look at them. All of this data should be reflected below in the graphs though.

```{r}
#| warning: FALSE
library(tidyverse)
library(readr)
library(readxl)
library(DT)
library(openxlsx)

spp <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/spp_models_diagnostics.xlsx"
genus <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/genus_models_diagnostics.xlsx"
fam <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/family_models_diagnostics.xlsx"
fg <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/fg_models_diagnostics.xlsx"
df <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/24and25predictions.xlsx"

spp_predictions <- read_excel(spp, "Sheet1")
genus_predictions <- read_excel(genus, "Sheet1")
fam_predictions <- read_excel(fam, "Sheet1")
fg_predictions <- read_excel(fg, "Sheet1")
predictions <- read_excel(df, "Sheet1")

pred <- predictions %>% 
  select(ID, Spp, pred_DryWeight, Percent, pred_source)

#left join this to covariate dbs
used_equations <- predictions %>% 
  distinct(Formula,.keep_all = TRUE) %>% 
  filter(!if_any(everything(), ~ grepl("observed", .x, ignore.case = TRUE))) %>% 
  select(-DryWeight, -pred_DryWeight, -Percent, -source_group, -ID, -PlotID)


datatable(
  pred,
  options = list(
    pageLength = 20,        # show more rows per page
    scrollX = TRUE,         # horizontal scrolling for wide tables
    scrollY = "400px",      # limit vertical size
    dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")  # center-align
    )
  ),
  class = "compact stripe hover",
  caption = "Predicted Biomass"
)

```

```{r}
#| warning: FALSE
#| eval: false
#| echo: false

#left join this to covariate dbs
used_equations <- predictions %>% 
  distinct(Formula, .keep_all = TRUE) 


low_r <- predictions %>% 
  distinct(R2, .keep_all = TRUE) %>% 
  filter(R2 < 0.6) %>% 
  select(Spp, Family, Genus, FunctionalGroup, pred_source, R2, Formula)

# datatable(
#   low_r,
#   options = list(
#     pageLength = 20,        # show more rows per page
#     scrollX = TRUE,         # horizontal scrolling for wide tables
#     scrollY = "400px",      # limit vertical size
#     dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
#     columnDefs = list(
#       list(className = 'dt-center', targets = "_all")  # center-align
#     )
#   ),
#   class = "compact stripe hover",
#   caption = "Low R-Squared"
# )


zeros <- predictions %>% 
 filter(Percent > 20 & pred_DryWeight == 0) %>% 
  select(ID, Spp, Family, Genus, FunctionalGroup, pred_DryWeight, Percent,pred_source, R2, Formula)



# datatable(
#   zeros,
#   options = list(
#     pageLength = 20,        # show more rows per page
#     scrollX = TRUE,         # horizontal scrolling for wide tables
#     scrollY = "400px",      # limit vertical size
#     dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
#     columnDefs = list(
#       list(className = 'dt-center', targets = "_all")  # center-align
#     )
#   ),
#   class = "compact stripe hover",
#   caption = "Flagged Predictions with Low Biomass"
# )
```

## Scatterplots of Predicted vs Observed Biomass

I went ahead and plotted each of the used equations to make sure I wasn't missing anything. The lists for each section highlight which equations have an R-squared value \<0.6. The stars next to any names indicate that it has both a low R-Squared value as well as it predicting low biomass for high percent cover in at least one instance. All of the graphs should contain the important information at the top, so you do not have to scroll back and forth as much.

### Species

-   BRTE_BROWN

-   BRTE_GREEN\*\*\*

-   ELEL5_GREEN

-   ERST4_ENTIRE

-   PHLO2_ENTIRE

-   POSE_BROWN\*\*\*

-   PSSP6_GREEN

-   VUMIM_BROWN

```{r}
library(ggplot2)
library(dplyr)
library(purrr)
library(glue)
library(patchwork)

spp_pred_species <- predictions %>%
  filter(pred_source == "Spp") %>%
  pull(Spp) %>%
  unique()

# 2. Keep Spp predictions and matching Observed rows
spp_predictions_filtered <- predictions %>%
  filter(
    pred_source == "Spp" |
      (pred_source == "Observed" & Spp %in% spp_pred_species)
  )%>% 
  arrange(Spp)

plot_species_predictions <- function(species_name, predictions, save_plots = FALSE) {
  
  # Filter for this species
  dat <- predictions %>% filter(Spp == species_name)
  
  # Prepare annotation text
  formula_text <- unique(dat$Formula[dat$source_group == "Predicted"])
  r2_text <- unique(dat$R2[dat$source_group == "Predicted"])
  pred_source_text <- unique(dat$pred_source[dat$source_group == "Predicted"])
  
  annotation_text <- if (length(formula_text) > 0) {
    glue("Source: {pred_source_text}\n{formula_text}\nR² = {r2_text}")
  } else {
    ""
  }
  
  # Base theme
  base_theme <- theme_bw(base_size = 12) +
    theme(
      legend.title = element_blank(),
      plot.subtitle = element_text(size = 10, face = "italic")
    )
  
  # Plot 1: pred_DryWeight vs Percent
  p1 <- ggplot(dat, aes(x = Percent, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      title = glue("{species_name} — Predicted Dry Weight"),
      subtitle = annotation_text,
      x = "Percent Cover",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Plot 2: pred_DryWeight vs ID
  p2 <- ggplot(dat, aes(x = ID, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      x = "Observation ID",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Combine plots side by side with legend at bottom
  combined_plot <- p1 + p2 + 
    plot_layout(guides = "collect") & 
    theme(legend.position = "bottom", legend.direction = "horizontal")
  
  # Save or display
  if (save_plots) {
    ggsave(glue("plots/{species_name}_pred_plot.png"), combined_plot, width = 10, height = 5, dpi = 300)
  } else {
    print(combined_plot)
  }
}



# Run for this subset
walk(unique(spp_predictions_filtered$Spp), ~plot_species_predictions(.x, spp_predictions_filtered))
```

### Genus 

-   ASTRAGULUS

-   BALSAMORHIZA

-   BROMUS

-   ERIOCOMA

-   ERIOGONUM

-   FESTUCA

-   LOMATIUM

-   POA\*\*\*\*

-   PSEUDOROEGNERIA\*\*\*

```{r}

genus_pred_species <- predictions %>%
  filter(pred_source == "Genus") %>%
  pull(Genus) %>%
  unique()

# 2. Keep Spp predictions and matching Observed rows
genus_predictions_filtered <- predictions %>%
  filter(
    pred_source == "Genus" |
      (pred_source == "Observed" & Genus %in% genus_pred_species)
  )%>% 
  arrange(Genus)

plot_genus_predictions <- function(genus_name, predictions, save_plots = FALSE) {
  
  # Filter for this species
  dat <- predictions %>% filter(Genus == genus_name)
  
  # Prepare annotation text
  formula_text <- unique(dat$Formula[dat$source_group == "Predicted"])
  r2_text <- unique(dat$R2[dat$source_group == "Predicted"])
  pred_source_text <- unique(dat$pred_source[dat$source_group == "Predicted"])
  
  annotation_text <- if (length(formula_text) > 0) {
    glue("Source: {pred_source_text}\n{formula_text}\nR² = {r2_text}")
  } else {
    ""
  }
  
  # Base theme
  base_theme <- theme_bw(base_size = 12) +
    theme(
      legend.title = element_blank(),
      plot.subtitle = element_text(size = 10, face = "italic")
    )
  
  # Plot 1: pred_DryWeight vs Percent
  p1 <- ggplot(dat, aes(x = Percent, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      title = glue("{genus_name} — Predicted Dry Weight"),
      subtitle = annotation_text,
      x = "Percent Cover",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Plot 2: pred_DryWeight vs ID
  p2 <- ggplot(dat, aes(x = ID, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      x = "Observation ID",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Combine plots side by side with legend at bottom
  combined_plot <- p1 + p2 + 
    plot_layout(guides = "collect") & 
    theme(legend.position = "bottom", legend.direction = "horizontal")
  
  # Save or display
  if (save_plots) {
    ggsave(glue("plots/{genus_name}_pred_plot.png"), combined_plot, width = 10, height = 5, dpi = 300)
  } else {
    print(combined_plot)
  }
}


walk(unique(genus_predictions_filtered$Genus), ~plot_genus_predictions(.x, genus_predictions_filtered))
```

### Family

-   APIACEAE

-   ASTERACEAE

-   POACEAE\*\*\*

```{r}
fam_pred_species <- predictions %>%
  filter(pred_source == "Family") %>%
  pull(Family) %>%
  unique()

# 2. Keep Spp predictions and matching Observed rows
fam_predictions_filtered <- predictions %>%
  filter(
    pred_source == "Family" |
      (pred_source == "Observed" & Family %in% fam_pred_species)
  )%>% 
  arrange(Family)

plot_fam_predictions <- function(fam_name, predictions, save_plots = FALSE) {
  
  # Filter for this species
  dat <- predictions %>% filter(Family == fam_name)
  
  # Prepare annotation text
  formula_text <- unique(dat$Formula[dat$source_group == "Predicted"])
  r2_text <- unique(dat$R2[dat$source_group == "Predicted"])
  pred_source_text <- unique(dat$pred_source[dat$source_group == "Predicted"])
  
  annotation_text <- if (length(formula_text) > 0) {
    glue("Source: {pred_source_text}\n{formula_text}\nR² = {r2_text}")
  } else {
    ""
  }
  
  # Base theme
  base_theme <- theme_bw(base_size = 12) +
    theme(
      legend.title = element_blank(),
      plot.subtitle = element_text(size = 10, face = "italic")
    )
  
  # Plot 1: pred_DryWeight vs Percent
  p1 <- ggplot(dat, aes(x = Percent, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      title = glue("{fam_name} — Predicted Dry Weight"),
      subtitle = annotation_text,
      x = "Percent Cover",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Plot 2: pred_DryWeight vs ID
  p2 <- ggplot(dat, aes(x = ID, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      x = "Observation ID",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Combine plots side by side with legend at bottom
  combined_plot <- p1 + p2 + 
    plot_layout(guides = "collect") & 
    theme(legend.position = "bottom", legend.direction = "horizontal")
  
  # Save or display
  if (save_plots) {
    ggsave(glue("plots/{fam_name}_pred_plot.png"), combined_plot, width = 10, height = 5, dpi = 300)
  } else {
    print(combined_plot)
  }
}


# Run for this subset
walk(unique(fam_predictions_filtered$Family), ~plot_fam_predictions(.x, fam_predictions_filtered))
```

### Functional Group

-   AF\*\*\*

-   PF

```{r}
#|warning: False
#|Message: False

fg_pred_species <- predictions %>%
  filter(pred_source == "FunctionalGroup") %>%
  pull(FunctionalGroup) %>%
  unique()

# 2. Keep Spp predictions and matching Observed rows
fg_predictions_filtered <- predictions %>%
  filter(
    pred_source == "FunctionalGroup" |
      (pred_source == "Observed" & FunctionalGroup %in% fg_pred_species)
  )%>% 
  arrange(FunctionalGroup)

plot_fg_predictions <- function(fg_name, predictions, save_plots = FALSE) {
  
  # Filter for this species
  dat <- predictions %>% filter(FunctionalGroup == fg_name)
  
  # Prepare annotation text
  formula_text <- unique(dat$Formula[dat$source_group == "Predicted"])
  r2_text <- unique(dat$R2[dat$source_group == "Predicted"])
  pred_source_text <- unique(dat$pred_source[dat$source_group == "Predicted"])
  
  annotation_text <- if (length(formula_text) > 0) {
    glue("Source: {pred_source_text}\n{formula_text}\nR² = {r2_text}")
  } else {
    ""
  }
  
  # Base theme
  base_theme <- theme_bw(base_size = 12) +
    theme(
      legend.title = element_blank(),
      plot.subtitle = element_text(size = 10, face = "italic")
    )
  
  # Plot 1: pred_DryWeight vs Percent
  p1 <- ggplot(dat, aes(x = Percent, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      title = glue("{fg_name} — Predicted Dry Weight"),
      subtitle = annotation_text,
      x = "Percent Cover",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Plot 2: pred_DryWeight vs ID
  p2 <- ggplot(dat, aes(x = ID, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    labs(
      x = "Observation ID",
      y = "Predicted Dry Weight (g)"
    ) +
    base_theme
  
  # Combine plots side by side with legend at bottom
  combined_plot <- p1 + p2 + 
    plot_layout(guides = "collect") & 
    theme(legend.position = "bottom", legend.direction = "horizontal")
  
  # Save or display
  if (save_plots) {
    ggsave(glue("plots/{fg_name}_pred_plot.png"), combined_plot, width = 10, height = 5, dpi = 300)
  } else {
    print(combined_plot)
  }
}

# Run for this subset
walk(unique(fg_predictions_filtered$FunctionalGroup), ~plot_fg_predictions(.x, fg_predictions_filtered))
```

## Regression Coefficients 

I am not sure if these are helpful, but we talked about looking at the coefficients previously so I thought I would include them just incase. Here are the R-squared, AIC, and Coeffecient values of the used regression equations that include an interaction term between various transformations of Julian Day and Percent cover. Ryan helped me code the interaction term into the dredging model, so every combination should have been considered.

### Species Level Regressions

```{r}
#| warning: FALSE

used_spp_predictions <- used_equations %>% 
  filter(pred_source == "Spp") %>%  
  select(Spp, R2, Formula) %>%
  left_join(spp_predictions, by = "R2") %>% 
  select(-Spp.y) %>% 
  rename(Spp = Spp.x) %>% 
  mutate(across(where(is.numeric), ~ round(.x, 4)))


datatable(
  used_spp_predictions,
  options = list(
    pageLength = 20,        # show more rows per page
    scrollX = TRUE,         # horizontal scrolling for wide tables
    scrollY = "400px",      # limit vertical size
    dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")  # center-align
    )
  ),
  class = "compact stripe hover",
  caption = "Species Coefficients"
)
```

### Genus Level Regressions

```{r}
used_genus_predictions <- used_equations %>% 
  filter(pred_source == "Genus") %>% 
  select(Genus, R2, Formula) %>% 
  left_join(genus_predictions, by = "R2") %>% 
  select(-Spp) %>% 
    mutate(across(where(is.numeric), ~ round(.x, 4)))

datatable(
  used_genus_predictions,
  options = list(
    pageLength = 20,        # show more rows per page
    scrollX = TRUE,         # horizontal scrolling for wide tables
    scrollY = "400px",      # limit vertical size
    dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")  # center-align
    )
  ),
  class = "compact stripe hover",
  caption = "Genus Coefficients"
)

```

### Family Level Regressions

```{r}
used_fam_predictions <- used_equations %>% 
  filter(pred_source == "Family") %>% 
  select(Family, R2, Formula) %>% 
  left_join(fam_predictions, by = "R2") %>% 
  select(-Spp) %>% 
    mutate(across(where(is.numeric), ~ round(.x, 4)))

datatable(
  used_fam_predictions,
  options = list(
    pageLength = 20,        # show more rows per page
    scrollX = TRUE,         # horizontal scrolling for wide tables
    scrollY = "400px",      # limit vertical size
    dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")  # center-align
    )
  ),
  class = "compact stripe hover",
  caption = "Familu Coefficients"
)

```

### FG Level Regressions 

```{r}
used_fg_predictions <- used_equations %>% 
  filter(pred_source == "FunctionalGroup") %>% 
  select(FunctionalGroup, R2, Formula) %>% 
  left_join(fg_predictions, by = "R2") %>% 
  select(-Spp) %>% 
    mutate(across(where(is.numeric), ~ round(.x, 4)))

datatable(
  used_fg_predictions,
  options = list(
    pageLength = 20,        # show more rows per page
    scrollX = TRUE,         # horizontal scrolling for wide tables
    scrollY = "400px",      # limit vertical size
    dom = "tip",            # minimal controls (t = table, i = info, p = pagination)
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")  # center-align
    )
  ),
  class = "compact stripe hover",
  caption = "Functional Group Coefficients"
)
```
