---
title: "Checking Bio Predict Assumptions"
author: "Alexis Means"
date: "2025-10-2"
description: This document contains assumption tests and plots to analyze my linear regression equations for biomass predictions. 
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
editor: visual
code-fold: true
---

```{r}
spp_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Species-Top-Model-List.rds")

fam_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Family-Top-Model-List.rds")

genus_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Genus-Top-Model-List.rds")

fg_model <- readRDS("C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/regression_equations/24and25-Biomass-Regression-Functional-Group-Top-Model-List.rds")

all_models <- list(
  spp   = spp_model,
  fam   = fam_model,
  genus = genus_model,
  fg    = fg_model
)

```

## Predicted Biomass Values

After including the interaction term between Julian Day and Percent cover in my linear regression models, here are the new predicted biomass outputs. I included a column that references which regression equation was used to predict the biomass. Relatively they are starting to look better, but there are still a lot of predicted zeros. The end of this documents includes scatterplots to better compare the predicted vs observed biomass values for each species.

These predictions do not include any of the species that were observed as 1% of cover in \>80% of the quadrats that the species was observed. I removed those to both create the linear regression equations as well as before predicting biomass for the remaining species. I followed Katey's procedure and predicted the biomass for those 1% species by taking the mean of the measured biomass values and upscaling it when needed. Currently I have the predictions using whichever equation ends up having the highest R2 value if none of the equations meet a 0.6 value which usually defaults back to the species level regression.

```{r}
#| warning: FALSE
library(readr)
library(readxl)
library(DT)

df <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/24and25predictions.xlsx"

predictions <- read_excel(df, "Sheet1")

datatable(predictions, options = list(pageLength = 10))
```

## R2, AIC and Coefficients Values

A lot of the species level regressions look a lot better and I believe are the main source of prediction. There are still a few that do not meet the R2 cutoff of 0.6. As you move up through the hierarchy of regression equations they progressively get worse. I am not sure what to do next to improve them. Or if it is worth working to improve them rather than just resulting in using the functional group like Katey did. However, most of the functional group regression equations currently have the worst R2 values.

### Species Level Regressions 

```{r}
spp <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/spp_models_diagnostics.xlsx"

spp_predictions <- read_excel(spp, "Sheet1")

datatable(spp_predictions, options = list(pageLength = 10))
```

### Genus Level Regressions 

```{r}
genus <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/genus_models_diagnostics.xlsx"

genus_predictions <- read_excel(genus, "Sheet1")

datatable(genus_predictions, options = list(pageLength = 10))

```

### Family Level Regressions

```{r}
fam <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/family_models_diagnostics.xlsx"

fam_predictions <- read_excel(fam, "Sheet1")

datatable(fam_predictions, options = list(pageLength = 10))
```

### Functional Group Regressions 

```{r}
fg <- "C:/Users/Alexis Means/Documents/Project/Nutrition Sampling/R code/biomass/processed.data/fg_models_diagnostics.xlsx"

fg_predictions <- read_excel(fg, "Sheet1")

datatable(fg_predictions, options = list(pageLength = 10))
```

```{r}
#ASSUMPTION TESTS
#| warning: FALSE
#| eval: FALSE
#| echo: FALSE
library(tidyverse)
library(lmtest)
library(purrr)
library(DT)

check_lm_assumptions <- function(model) {
  if (!inherits(model, "lm")) return(NULL)
  
  res <- residuals(model)
  
  list(
    r_squared = tryCatch(round(summary(model)$r.squared, 4), error = function(e) NA),
    shapiro_p = tryCatch(round(shapiro.test(res)$p.value, 4), error = function(e) NA),
    bp_test   = tryCatch(round(lmtest::bptest(model)$p.value, 4), error = function(e) NA),
    dw_test   = tryCatch(round(lmtest::dwtest(model)$p.value, 4), error = function(e) NA),
    cook_max  = tryCatch(round(max(cooks.distance(model), na.rm = TRUE), 4), error = function(e) NA)
  )
}

diagnostics <- imap_dfr(all_models, function(model_list, group_name) {
  map_dfr(model_list, check_lm_assumptions, .id = "model_name") %>%
    mutate(group = group_name, .before = 1)
})

#--------------------------------------------------------------
# 3️⃣ Extract model base name (everything before trailing number)
#     Example: "BRTE_GREEN1" → "BRTE_GREEN"
#--------------------------------------------------------------
diagnostics <- diagnostics %>%
  mutate(model_base = str_trim(str_extract(model_name, ".*(?=\\d+$)")))

#--------------------------------------------------------------
# 4️⃣ Identify and remove model families with any R² > 0.4
#--------------------------------------------------------------
high_r2_bases <- diagnostics %>%
  group_by(model_base) %>%
  summarize(any_high_r2 = any(r_squared > 0.4, na.rm = TRUE)) %>%
  filter(any_high_r2) %>%
  pull(model_base)

diagnostics_low_r2 <- diagnostics %>%
  filter(!model_base %in% high_r2_bases)


datatable(diagnostics_low_r2, options = list(pageLength = 10))
```

## Scatterplots of Predicted vs Observed Biomass

Sorry in advance for the extensive amounts of scatterplots below. I plotted the biomass measurement based on observation ID as well as Percent cover in order to compare the predicted vs observed biomass values. The plot with the observation ID is meant more to flag any points that I may need to remove in order to help the predictions.

```{r}
#| message: false
#| warning: false
#| fig.height: 4
#| fig.width: 10
#| fig.align: center


library(patchwork)  # <-- important for combining plots

# Create color grouping
predictions <- predictions %>%
  mutate(source_group = ifelse(pred_source == "Observed", "Observed", "Predicted"))

# Get unique species list
species_list <- unique(predictions$Spp)

# Loop through species and print each combined plot
for (sp in species_list) {
  dat <- filter(predictions, Spp == sp)
  
  # Plot 1: x = ID
  p1 <- ggplot(dat, aes(x = factor(ID), y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "steelblue", "Predicted" = "orange")) +
    labs(
      title = "By Observation ID",
      x = "Observation ID",
      y = "Predicted Dry Weight (g)",
      color = "Data Source"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      legend.position = "top",
      legend.direction = "vertical",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  # Plot 2: x = Percent
  p2 <- ggplot(dat, aes(x = Percent, y = pred_DryWeight, color = source_group)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "steelblue", "Predicted" = "orange")) +
    labs(
      title = "By Percent Cover",
      x = "Percent",
      y = "Predicted Dry Weight (g)",
      color = "Data Source"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      legend.position = "top",
      legend.direction = "vertical"
    )
  
  # Combine side-by-side and share legend vertically above both plots
  combined <- (p1 + p2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "top", legend.direction = "vertical")
  
  # Add species name as a single annotation above everything
  final_plot <- combined +
    plot_annotation(
      title = paste("Predicted Dry Weight for", sp),
      theme = theme(
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
      )
    )
  
  print(final_plot)
  cat("\n\n")
}
```

## What's next?

-   removing points and rerunning regression equations?

-   restructuring my functional groups to be more specific about tall vs short forbs and things like that.
